Application of Network Flow and Linear Programming to Scheduling of Clinicians

### 2019-07-19 paper/paper.methods.tex
Minor fixes to problem, methods sections

\subsection{Variables}
Since each clinician may be assigned to work for any service, during any block of the year, we denote such an assignment as $X_{c, b, d}$. A value of 1 indicates that the given clinician $c$ is assigned to cover division $d$ during block $b$. Similarly, we define weekend assignments as $Y_{c, w}$, although these are not indexed by service, as clinicians are expected to provide all services during the weekends. The optimization of the soft constraint Block-Weekend Adjacency is done by maximizing the product $X_{c, b, d} \cdot Y_{c, w}$ for adjacent blocks and weekends. In order to formulate such an objective as a linear function of variables, we introduce another set of variables, denoted by $Z_{c, b, d}$, with additional constraints on its range. Further details regarding this variable are described in Sections \ref{???} and \ref{???}. Lastly, we introduce a set of constants $m_{c, d}$ and $M_{c, d}$ to constrain the number of blocks each clinician is allowed to work during the year. Table \ref{tbl:variables-constants} presents a summary of the constants and variables in the problem.

@@

\subsection{Objectives}
As mentioned briefly in Section \ref{???}, the soft constraints of the clinician scheduling problem are: (1) satisfying clinician block off requests, (2) satisfying clinician weekend off requests, and (3) assigning weekends closer to blocks. We convert these constraints into linear objective functions of the binary variables defined in Section \ref{???}. The first two objectives have a straight-forward formulation as linear functions:
\begin{align}
	&Q_1(X) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{d=1}^{D} (-1)^{\ind(b\,\in\,\mathcal{BR}_c)}\cdot X_{c, b, d} \tag{Block Requests} \label{eqn:obj-block-requests}\\
	&Q_2(Y) = \sum_{c=1}^{C} \sum_{w=1}^{W} (-1)^{\ind(w\,\in\,\mathcal{WR}_c)}\cdot Y_{c, w} \tag{Weekend Requests} \label{eqn:obj-weekend-requests}
\end{align}
where $\ind(\cdot)$ is the indicator function that has value 1 when the condition is met and 0 otherwise. In these two objectives, we penalize any assignments that conflict with a block or weekend request, and aim to maximize the non-conflicting assignments. The Block-Weekend Adjacency is optimized by considering the product $X_{c, b, d}\cdot Y_{c, w}$ for values of $w$ ``adjacent'' to the value of $b$. For instance, clinicians might want to be assigned during a weekend that falls within an assigned block. So if clinician $c$ is assigned to work during block 3, corresponding to weeks 5 and 6 assuming 2-week blocks, they might also want to be assigned to work during weekend 5. In that case, we would like $X_{c, b=3, d} \cdot Y_{c, w=5}$ to be 1, since that indicates both variables are assigned. If at least one of the two variables is not assigned, the product will be 0. This leads to the obvious maximization objective
\begin{align}
	&Q_3(X, Y) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{d=1}^{D} X_{c, b, d}\cdot Y_{c, w=\varphi(b)} \tag{Block-Weekend Adjacency} \label{eqn:obj-block-weekend-adj}
\end{align}
where $\varphi(b)$ maps a block one-to-one to an adjacent weekend, by some appropriate definition of adjacency. For instance, in the above example we will have $\varphi(b) = 2b - 1$. \\

>>>>

\subsection{Variables}
Since each clinician may be assigned to work for any service, during any block of the year, we denote such an assignment as $X_{c, b, d}$. A value of 1 indicates that the given clinician $c$ is assigned to cover division $d$ during block $b$. Similarly, we define weekend assignments as $Y_{c, w}$, although these are not indexed by service, as clinicians are expected to provide all services during the weekends. The optimization of the soft constraint Block-Weekend Adjacency is done by maximizing the product $X_{c, b, d} \cdot Y_{c, w}$ for adjacent blocks and weekends. In order to formulate such an objective as a linear function of variables, we introduce another set of variables, denoted by $Z_{c, b, d}$, with additional constraints on its range. Further details regarding this variable are described in Sections \ref{???} and \ref{???}. Lastly, we introduce a set of constants $m_{c, d}$ and $M_{c, d}$ to constrain the minimal and maximal number of blocks each clinician is allowed to work during the year. Table \ref{tbl:variables-constants} presents a summary of the constants and variables in the problem.

@@

\subsection{Objectives}
As mentioned briefly in Section \ref{???}, the soft constraints of the clinician scheduling problem are: (1) satisfying clinician block off requests, (2) satisfying clinician weekend off requests, and (3) assigning weekends closer to blocks. We convert these constraints into linear objective functions of the binary variables defined in Section \ref{???}. The first two objectives have a straightforward formulation as linear functions:
\begin{align}
	&Q_1(X) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{d=1}^{D} (-1)^{\ind(b\,\in\,\mathcal{BR}_c)}\cdot X_{c, b, d} \tag{Block Requests} \label{eqn:obj-block-requests}\\
	&Q_2(Y) = \sum_{c=1}^{C} \sum_{w=1}^{W} (-1)^{\ind(w\,\in\,\mathcal{WR}_c)}\cdot Y_{c, w} \tag{Weekend Requests} \label{eqn:obj-weekend-requests}
\end{align}
where $\ind(\cdot)$ is the indicator function that has value 1 when the condition is met and 0 otherwise. In these two objectives, we penalize any assignments that conflict with a block or weekend request, and aim to maximize the non-conflicting assignments. The Block-Weekend Adjacency is optimized by considering the product $X_{c, b, d}\cdot Y_{c, w}$ for values of $w$ ``adjacent'' to the value of $b$. For instance, clinicians might want to be assigned during a weekend that falls within an assigned block. So if clinician $c$ is assigned to work during block 3, corresponding to weeks 5 and 6 assuming 2-week blocks, they might also want to be assigned to work during weekend 5. In that case, we would like $X_{c, b=3, d} \cdot Y_{c, w=5}$ to be 1, since that indicates both variables are assigned. If at least one of the two variables is not assigned, the product will be 0. This leads to the obvious maximization objective
\begin{align}
	&Q_3(X, Y) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{d=1}^{D} X_{c, b, d}\cdot Y_{c, w=\varphi(b)} \tag{Block-Weekend Adjacency} \label{eqn:obj-block-weekend-adj}
\end{align}
where $\varphi(b)$ maps a block one-to-one to an adjacent weekend, by some appropriate definition of adjacency. For instance, in the above example we will have $\varphi(b) = 2b - 1$. \\

### 2019-07-19 paper/paper.problem.tex
Minor fixes to problem, methods sections

Several constraints are placed on the on-call assignments. First, each clinician has limits on the number of blocks they can and must work during the year, depending on the type of consultation. For instance, a clinician might have to provide 3-5 blocks of general ID consultation as well as 2-3 blocks of HIV consultation throughout the year. These limits may change from year to year as the number of clinicians in the department changes. Moreover, the schedule should not assign a clinician to work for two blocks or two weekends in a row. In addition, the schedule should distribute both regular and holiday weekends equally among all clinicians. \\

@@

\begin{table}[h]
	\centering
	\begin{tabular}{ l l }
		\hline
		\textbf{Constraint Name} & \textbf{Description}                                                                                                           \\ \hline
		Block Coverage           & \makecell[l]{each division needs to have exactly \\ one clinician that covers any given block}                                 \\ \hline
		Weekend Coverage         & \makecell[l]{every weekend needs to have exactly \\ one clinician that covers it}                                              \\ \hline
		Min/Max                  & \makecell[l]{in a given division, each clinician can only \\ work between the minimum and maximum \\ number of allowed blocks} \\ \hline
		No Consecutive Blocks    & \makecell[l]{any clinician should not work \\ two consecutive blocks, across all divisions}                                    \\ \hline
		No Consecutive Weekends  & \makecell[l]{any clinician should not work two consecutive \\ weekends}                                                        \\ \hline
		Equal Weekends           & \makecell[l]{weekends should be equally distributed \\ between clinicians}                                                     \\ \hline
		Equal Holidays           & \makecell[l]{long weekends should be equally distributed \\ between clinicians}                                                \\ \hline
		Block Requests           & \makecell[l]{each clinician can request to be off service \\ during certain blocks throughout the year}                        \\ \hline
		Weekend Requests         & \makecell[l]{each clinician can request to be off service \\ during certain weekends throughout the year}                      \\ \hline
		Block-Weekend Adjacency  & \makecell[l]{the block and weekend assignments of a given \\ clinician should be adjacent}
	\end{tabular}
	\caption{Summary of the constraints for the clinician scheduling problem}
	\label{tbl:constraint-summary}
\end{table}

In most scheduling problems, the constraints can be divided into hard and soft constraints. Hard constraints must be satisfied by any candidate solution, while soft constraints can be used to select a more favourable solution from all candidate solutions [ref]. Typically, soft constraints are encoded as an objective that needs to be maximized or minimized, rather than a constraint. In the case of the clinician scheduling problem, we chose Block Requests, Weekend Requests and Block-Weekend Adjacency as our soft constraints, while the rest of the constraints are hard. Though it is important to take clinician requests into account when constructing the schedule, it is crucial to that the work-load of the schedule is balanced among all clinicians, and the needs of the patients are fulfilled. \\

>>>>

Several constraints are placed on the on-call assignments. First, each clinician has limits on the number of blocks they can and must work during the year, depending on the type of consultation. For instance, a clinician might have to provide 3-5 blocks of general ID consultation as well as 2-3 blocks of HIV consultation throughout the year. These limits may change from year to year as the number of clinicians in the department changes. Moreover, the schedule should not assign a clinician to work for two blocks or two weekends in a row. The schedule should also distribute both regular and holiday weekends equally among all clinicians. \\

@@

\begin{table}[h]
	\centering
	\begin{tabular}{ l l l }
		\hline
		\textbf{Constraint Name} & \textbf{Description}                                                                                                           & \textbf{Type} \\ \hline
		Block Coverage           & \makecell[l]{each division needs to have exactly \\ one clinician that covers any given block}                                 & Hard          \\ \hline
		Weekend Coverage         & \makecell[l]{every weekend needs to have exactly \\ one clinician that covers it}                                              & Hard          \\ \hline
		Min/Max                  & \makecell[l]{in a given division, each clinician can only \\ work between the minimum and maximum \\ number of allowed blocks} & Hard          \\ \hline
		No Consecutive Blocks    & \makecell[l]{any clinician should not work \\ two consecutive blocks, across all divisions}                                    & Hard          \\ \hline
		No Consecutive Weekends  & \makecell[l]{any clinician should not work two consecutive \\ weekends}                                                        & Hard          \\ \hline
		Equal Weekends           & \makecell[l]{weekends should be equally distributed \\ between clinicians}                                                     & Hard          \\ \hline
		Equal Holidays           & \makecell[l]{long weekends should be equally distributed \\ between clinicians}                                                & Hard          \\ \hline
		Block Requests           & \makecell[l]{each clinician can request to be off service \\ during certain blocks throughout the year}                        & Soft          \\ \hline
		Weekend Requests         & \makecell[l]{each clinician can request to be off service \\ during certain weekends throughout the year}                      & Soft          \\ \hline
		Block-Weekend Adjacency  & \makecell[l]{the block and weekend assignments of a given \\ clinician should be adjacent}                                     & Soft
	\end{tabular}
	\caption{Summary of the constraints for the clinician scheduling problem}
	\label{tbl:constraint-summary}
\end{table}

In most scheduling problems, the constraints can be divided into hard and soft constraints. Hard constraints must be satisfied by any candidate solution, while soft constraints can be used to select a more favourable solution from all candidate solutions [ref]. Typically, soft constraints are encoded as objective functions rather than actual constraints. These objective functions are maximized or minimized when finding a solution. In the case of our clinician scheduling problem, we chose Block Requests, Weekend Requests and Block-Weekend Adjacency as our soft constraints, while the rest of the constraints are hard. Though it is important to take clinician requests into account when constructing the schedule, it is crucial that the work-load of the schedule is balanced among all clinicians, and the needs of the patients are fulfilled. \\
